---
title: "Homework #10"
author: "Nicole R Gorman"
date: "5 April 2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#For loops and randomization tests  

* Open libraries
```{r open libraries, message=FALSE}
library(tidyverse)
library(ggplot2)
library(MASS)
library(gridExtra)
```

```
# Let's talk about tea
# Is milk added first or not?
# H0 vs HA
# Blind test
# Can reshuffle the data to repeat the experiment rather than actually redoing the experiment
```

1. Using a for loop, write a function to calculate the number of zeroes in a numeric vector. Before entering the loop, set up a counter variable counter <- 0. Inside the loop, add 1 to counter each time you have a zero in the vector. Finally, use return(counter) for the output.

```{r echo=TRUE}

vec <- rbinom (n=5, size=1, 0.5)
counter <- 0
# create a vector w/ 0's and set up a counter variable

##########################################################################
# FUNCTION count_zeros
# count the number of zeros in a vector
# input: vector with zeros
# output: number of variables = 0
#----------------------------------------------------------------

count_zeros <- function(vec) {
 for (i in vec) {
   if(i==0) {
   counter <- counter + 1
}
   return(counter)
}
}

# end of count_zeros function
##########################################################################

count_zeros(vec)

```

2. Use subsetting instead of a loop to rewrite the function as a single line of code.

```{r echo=TRUE, results=TRUE}

vec <- rbinom (n=5, size=1, 0.5)
counter <- 0

##########################################################################
# FUNCTION count_zeros2
# count the number of zeros in a vector
# input: vector with zeros
# output: number of variables = 0
#----------------------------------------------------------------

count_zeros2 <- function(vec) {
   return(sum(vec==0))
}

# end of count_zeros2 function
##########################################################################

result <- count_zeros2(vec)
print (result)
# call function and print result
```


3. Write a function that takes as input two integers representing the number of rows and columns in a matrix. The output is a matrix of these dimensions in which each element is the product of the row number x the column number.

```{r echo=TRUE}

##########################################################################
# FUNCTION prod_matrix
# create matrix with elements as products of row and column
# input: number of rows and columns
# output: matrix of products
#----------------------------------------------------------------

prod_matrix <- function(nRow,nCol) {
   my_mat <- matrix(0, nrow = nRow, ncol = nCol)
  
  for (i in 1:nRow) {
    for (j in 1:nCol) {
       my_mat[i,j] <- i * j
  }
  
}
  return(my_mat)
}

# end of prod_matrix function
##########################################################################

my_matrix <- prod_matrix(3,4)
print (my_matrix)

```

4. Now let’s practice calling custom functions within a for loops. Use the code from previous lectures on loops and functions to complete the following steps:

a. Simulate a dataset with 3 groups of data, each group drawn from a distribution with a different mean. The final data frame should have 1 column for group and 1 column for the response variable.

```{r echo=TRUE}

# Set seed for reproducibility
set.seed(1)

# Simulate data for each group
group1 <- rnorm(50, mean = 10, sd = 1)
group2 <- rnorm(50, mean = 15, sd = 1)
group3 <- rnorm(50, mean = 20, sd = 1)

# Combine data into a single dataframe
data <- data.frame(
  group = rep(c("Group1", "Group2", "Group3"), each = 50),
  response = c(group1, group2, group3)
)

head(data)
tail(data)

```

```{r echo=TRUE}

# Convert mean values to numeric if necessary
# group1 <- as.numeric(group1)
# group2 <- as.numeric(group2)
# group3 <- as.numeric(group3)
# 
# # Create histograms of the means for each group using qplot
# Ori_Group1 <- qplot(data = data, x = group1, geom = "histogram", binwidth = 1, alpha = I(0.3),fill = "green") +
#   labs(title = "Original Mean Group1", x = "Mean", y = "Frequency")
# 
# Ori_Group2 <- qplot(data = data, x = group2, geom = "histogram", binwidth = 1, alpha = I(0.3), fill = "blue") +
#   labs(title = "Original Mean Group2", x = "Mean", y = "Frequency")
# 
# Ori_Group3 <- qplot(data = data, x = group3, geom = "histogram", binwidth = 1, alpha = I(0.3), fill = "red") +
#   labs(title = "Original Mean Group3", x = "Mean", y = "Frequency")
# 
# # Arrange histograms
# grid.arrange(Ori_Group1, Ori_Group2, Ori_Group3, ncol = 3)
```


b. Write a custom function that 1) reshuffles the response variable, and 2) calculates the mean of each group in the reshuffled data. Store the means in a vector of length 3.

```{r echo=TRUE}
##########################################################################
# FUNCTION shuffle_mean
# reshuffle response variable, calculate the mean of each group, store the means in a vector of length 3.
# input: dataset with 3 groups
# output: three group means as vector
#----------------------------------------------------------------

shuffle_mean <- function(data, resVar, groupVar) {
   
  # reshuffle response variable
  resVar2 <- sample(data[[resVar]])
   
  # calculate the mean of each group in data2
  meanGroup <- tapply(resVar2, data[[groupVar]], mean)
  
  # Convert mean_group to a vector of length 3
  meanVector <- numeric(3)
  for (i in 1:3) {
    meanVector[i] <- meanGroup[[i]]
  }
  
  # Return the vector of means
  return(meanVector)
}

# end of shuffle_mean function
##########################################################################

meanVector <- shuffle_mean(data, "response", "group")

# Print the vector of means
print(meanVector)

```


c. Use a for loop to repeat the function in b 100 times. Store the results in a data frame that has 1 column indicating the replicate number and 1 column for each new group mean, for a total of 4 columns.

```{r echo=TRUE}

# Create an empty data frame to store results
# z <- rep(1:100)
df <- data.frame(repNum = 1:100, Group1 = numeric(100), Group2 = numeric(100), Group3 = numeric(100))

# Repeat the function 100 times
for (i in 1:100) {
  
  # Call function
  meanVector <- shuffle_mean(data, "response", "group")
  
  # Add the means to the data frame
  df[i, -1] <- meanVector
}

# Rename columns
colnames(df)[-1] <- paste0("Group", 1:3)

# Print the results data frame
print(df)

```


d. Use qplot() to create a histogram of the means for each reshuffled group. Or, if you want a challenge, use ggplot() to overlay all 3 histograms in the same figure. 

```{r echo=TRUE}

# Create histograms of the means for each group using qplot
hist_Group1 <- qplot(data = df, x = Group1, geom = "histogram", cex = .3, binwidth = 1, alpha = I(0.3),fill = "green") +
  labs(title = "Shuffled Mean Group1", x = "Mean", y = "Frequency")

hist_Group2 <- qplot(data = df, x = Group2, geom = "histogram", cex = .3, binwidth = 1, alpha = I(0.3), fill = "blue") +
  labs(title = "Shuffled Mean Group2", x = "Mean", y = "Frequency")

hist_Group3 <- qplot(data = df, x = Group3, geom = "histogram", cex = .3, binwidth = 1, alpha = I(0.3), fill = "red") +
  labs(title = "Shuffled Mean Group3", x = "Mean", y = "Frequency")

# Arrange histograms
grid.arrange(hist_Group1, hist_Group2, hist_Group3, ncol = 3)
```

How do the distributions of reshuffled means compare to the original means?

```{r echo=TRUE}
# Still working oin the answer to this question...
```