---
title: "Homework9"
author: "Nicole R Gorman"
date: "2024-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Organizing Code With Structured Programming

* Open libraries
```{r openLibraries, message=FALSE}
library(tidyverse)
library(ggplot2)
library(MASS)
library(dplyr)
```

* Use the code that you worked on in Homework #7 (creating fake data sets), and re-organize it following the principles of structured programming.  
* Do all the work in a single chunk in your R markdown file, just as if you were writing a single R script. 
* Start with all of your annotated functions, preliminary calls, and global variables. 
* The program body should be only a few lines of code that call the appropriate functions and run them in the correct order. 
* Make sure that the output from one function serves as the input to the next. You can either daisy-chain the functions or write separate lines of code to hold elements in temporary variables and pass them along.

```{r readInData, warnings=FALSE, echo=TRUE, results='hide'}
# Read in the data
setwd("~/Documents/UVM_Research/UVM Rotation Project/ACC Trials")
x<-list.files(pattern="HW9")

```

```{r cln_dat, cln_dat, warnings=FALSE, echo=TRUE, results=TRUE}

# call the new functions with my data set as an argument
source("Homework9Functions.R")

my_data <- cln_dat("ACCData_HW9.csv", "genotype", "treatment", "length")
# returns clean data set
# not working correctly, renaming individual samples with generic names in dataframe. I must have something flipped?
```


```{r dat_stat, echo=TRUE, results=TRUE}

result <- dat_stat(my_data,"resVar")
print(result)
# returns histogram and summary statistics 

```

```{r ANOVA, echo=TRUE, results=TRUE}

ANOplot <- anova(my_data, "resVar", 6, c("WT", "CCDC22", "CCDC93", "CCDC22CCDC93", "CCDC22RFP", "CCDC93RFP"))
print(ANOplot)
# returns anova results
# not working correctly, cannot get the mean of each group(genotype), calculating the mean of the entire dataset instead
```

**I did not get to this second part...still working on correcting the errors in question #1**

* Once your code is up and working, modify your program to do something else: record a new summary variable, code a new statistical analysis, or create a different set of random variables or output graph. 
* Do not rewrite any of your existing functions. Instead, copy them, rename them, and then modify them to do new things. 
* Once your new functions are written, add some more lines of program code, calling a mixture of your previous functions and your new functions to get the job done.

```{r echo=TRUE, results=TRUE}


# cleaned_data2 <- cln_dat(z) 
# # returns clean dataset with no NAs
# 
# gen_data2 <- my_vars(cleaned_data)
# # returns dataset with generic variable names 
# 
# dat_stats2 <- dat_stat(gen_data)
# # plots dataset and returns stats mean and sd

```


* Optional. If time permits and you have the skills, try putting your program inside of a for loop and repeat the analysis with a different stochastic data set 
# (each time you call a function that invokes the random number generator, it will create a new set of data for you to process). 
* Can you create a data structure to store the summary statistics created in each pass through the loop? 
* If not, your program will work, but it will only show the results from the final replicate (the previous results will be written over each time you traverse the loop).

```{r echo=TRUE, results=TRUE}
 

```
