---
title: "Homework #10"
author: "Nicole R Gorman"
date: "5 April 2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```
#For loops and randomization tests  

* Open libraries
```{r open libraries, message=FALSE}
library(tidyverse)
library(ggplot2)
library(MASS)
```

1. Using a for loop, write a function to calculate the number of zeroes in a numeric vector. Before entering the loop, set up a counter variable counter <- 0. Inside the loop, add 1 to counter each time you have a zero in the vector. Finally, use return(counter) for the output.

```{r echo=TRUE}

z <- rbinom(n=5, size=1, 0.5)

counter <- 0

count_zeros <- function(vec) {
 for (i in seq_along(z)) {
 if(z[i]==0) {
  counter <- i + 1
}
   return(counter)
}
}

count_zeros(z)

```

2. Use subsetting instead of a loop to rewrite the function as a single line of code.

```{r echo=TRUE}

z <- rbinom(n=5, size=1, 0.5)
z <- if(z[i]==0), z + 1)
print(z)

```



Write a function that takes as input two integers representing the number of rows and columns in a matrix. The output is a matrix of these dimensions in which each element is the product of the row number x the column number.

Now let’s practice calling custom functions within a for loops. Use the code from previous lectures on loops and functions to complete the following steps:

Simulate a dataset with 3 groups of data, each group drawn from a distribution with a different mean. The final data frame should have 1 column for group and 1 column for the response variable.

Write a custom function that 1) reshuffles the response variable, and 2) calculates the mean of each group in the reshuffled data. Store the means in a vector of length 3.

Use a for loop to repeat the function in b 100 times. Store the results in a data frame that has 1 column indicating the replicate number and 1 column for each new group mean, for a total of 4 columns.

Use qplot() to create a histogram of the means for each reshuffled group. Or, if you want a challenge, use ggplot() to overlay all 3 histograms in the same figure. How do the distributions of reshuffled means compare to the original means?


```
#Let's talk about tea
# Is milk added first or not?
# H0 vs HA
# Blind test
# can reshuffle the data to repeat the experiment rather than actually redoing the experiment
```

```
# question 3 is going to require double for loop
```
