---
title: "Homework12"
author: "Nicole R Gorman"
date: "2024-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* Load libraries

```{r open libraries, message=FALSE}
library(dplyr)
library(ggplot2)
library(ggthemes)
library(patchwork)
```

```
For this exercise, use your newly-developed ggplot chops to create some nice graphs from your own data (If you do not have a good data frame to use for graphics, use one of the many built-in data frames from R (other than mpg, which we are using in class)). Experiment with different themes, theme base sizes, aesthetics, mappings, and faceting. When you are finished, try exporting them to high quality pdfs, jpgs, eps files, or other formats that you would use for submission to a journal.
```
**Components of a ggplot layer**

```
1. data frame: load data, create dataframe
2. variables mapped to aesthetics
3. geom = geometric used to draw the layer
4. stat = takes raw data and transforms it for useful plotting
5. position = adjustment prevents overplotting
```

```{r echo=TRUE, results=FALSE}
# May 4th demands use of built in starwars data frame
# read in/load data

data(starwars) 
data <- data.frame(starwars) 
```


```{r echo=TRUE, results=FALSE}
# explore the data just a bit to see what might be interesting to plot

str(data)
print(data)
table(data$height)
table(data$mass) 
table(data$species)
```

* It looks like I need to create some categories and/or summary statistics of the data for plotting. I'll try that in a bit. For now, I'll take a quick look at a basic histogram looking at species.

```{r echo=TRUE, results=TRUE}
p1 <- ggplot(data=data) +
      aes(x=species) +
      geom_bar(fill="khaki",color="black") +
      labs(title="Wow! This Graph Is Terrible",
          subtitle="Let's see what I can do to fix it...",
          x="Species",
          y="Character Count")
print(p1)
```      

* I can use a summary statistic to make some sense of the data and also add some elements to the plot for a better visual summmary of the data. Add text annotations to a graph.

```{r echo=TRUE, results=TRUE}

# First I need to filter out missing values
starwars <- starwars %>% filter(!is.na(height))

# Now some summary stats
# Calculate the mean height standard error (insufficient data to find standard deviation) for each species 
species_height <- starwars %>%
  group_by(species) %>%
  summarize(avg_height = mean(height, na.rm = TRUE),
            se_height = sd(height, na.rm = TRUE) / sqrt(n()))

# Sort the data by average height
species_height <- species_height[order(species_height$avg_height), ]

# Create the bar plot to visualize the relationship of height and species
p2 <- ggplot(species_height, aes(x = species, y = avg_height)) +
      geom_bar(stat = "identity", fill = "skyblue") +
      geom_errorbar(aes(ymin = avg_height - se_height, ymax = avg_height + 
                    se_height), width = 0.3, color = "blue") + 
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(face = "bold", size = 14), 
      axis.title.x = element_text(face = "bold", size = 11),  
      axis.title.y = element_text(face = "bold", size = 11)) + 
      labs(x = "Species", y = "Mean Height (cm)", 
           title = "Mean Height of Star Wars Character Species",
           subtitle="Still could use some help, but much better!")
print(p2)

# I tried to add a note at the bottom to point out interesting data, but I was not able to adjust the height appropriately so it would not interfere with the bar plot. It is commented out below.

# p1 + annotate("text", x = Inf, y = -Inf, hjust = 1, vjust = -.75, 
#             label = "*Note relatively large error bar for droids may reflect phenotypic variability or small sample size", size = 3) +
#  theme(plot.margin = margin(b = -1.1, unit = "lines"))

```

* Compare the plots

```{r echo=TRUE, results=TRUE}
# Place the plots vertically for easier comparison
  p1 + p2 + plot_layout(ncol=1)
```

* Explore another type of geom

```{r echo=TRUE, results=TRUE}

# I have to convert 'homeworld' to numeric in order to show a trend line
data$homeworld_numeric <- as.numeric(factor(data$homeworld))

# Check the structure of the data 
# str(data)

# Create a jitter dot plot
# Make dots transparent to visualize overlap/clustering

p3 <- ggplot(data = data, aes(x = homeworld, y = species)) +
      geom_jitter(alpha = 0.35, color = "blue", size = 4) +
       labs(x = "Homeworld", y = "Species", 
       title = "Identification of Species Adaptive Radiation Potential 
       Based on Prevalence in Multiple Homeworlds") +
       theme(plot.title = element_text(size = 17,hjust = .3))  
       # Change title size and center 

# Rotate x-axis labels by 45 degrees
p3 <- p3 + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
                 axis.text.y = element_text(size = 7))

print(p3)
```

Change theme for a more engaging visual. Spacig between labels and title still needs to be adjusted, but theme improves visual impact.

```{r echo=TRUE, results=TRUE}

# histogram
p4 <- ggplot(data=data) +
       aes(x=homeworld) +
       geom_bar (fill="khaki",color="black")
       
# many specialized themes
p4 + theme_economist() +
     theme(axis.text.x = element_text(angle = 45, hjust = .05, vjust = .08, size = 7, margin = margin(0, 0, 0, 0.05, "cm"))) +
     labs(title="Patterns of Distribution of Star Wars Characters",
         subtitle="Not sure what I could say here, but something to help clarify the purpose of this plot",
         x="Homeworld",
         y="Number of Characters") 
```
 
* I tried to look at transportation, but couldn't quite get my mind around what might be interesting to compare. 

```{r echo=TRUE, results=TRUE} 

# Get unique vehicles
unique_vehicles <- unique(unlist(starwars$vehicles))

# Create a summary data frame with homeworld and number of vehicles
homeworld_vehicles <- starwars %>%
  group_by(homeworld) %>%
  summarize(total_vehicles = sum(!is.na(vehicles)))

# None of the plots I tried showed anything worth looking at, but it seems like there should be something here.
# 
#     labs(title = "The Relative Frequency of Motorized Transportation by Homeworld",
#     subtitle = "An extended subtitle that will print below the main title",
#            x = "Homeworld",
#            y = "Total Vehicles") 

# Print the plot
# print(p5)
```

* Change the labels on a facet plot so that they are more readable informative than the variable names that are supplied from your data frame. 
* Note that figure shows a trend that other life forms are more dense than both humans and droids.

``` {r echo=TRUE, results=TRUE} 
 
# Filter out missing values
starwars <- starwars %>% filter(!is.na(height) & !is.na(mass))

# Not a great option because the other category need additional sorting criteria
# However, with additional sorting this could be a valuable tool for analysis

# Change "Unknown" to "Other life forms" in the species column
starwars <- starwars %>% 
  mutate(species = case_when(
    species == "Unknown" ~ "Other Life Forms",
    TRUE ~ species
  ))

# Categorize species to try and identify a relationship between mass and height related to species
starwars <- starwars %>% 
  mutate(species = ifelse(species == "", "Unknown", species)) %>% 
  mutate(species_group = ifelse(species %in% c("Human", "Droid", "Other Life Forms"), species, "Other"))

# Create the facet plot
ggplot(starwars, aes(x = height, y = mass)) +
  geom_point(alpha = 0.5, color = "darkgreen", size =5) +
  facet_wrap(~ species_group, scales = "free") +
  labs(x = "Height", y = "Mass", title = "Height vs. Mass by Species Group") +
  theme(plot.title = element_text(size = 20, hjust = .5))  

```



