---
title: "Homework #4: Atomic Vectors"
author: "Nicole R Gorman"
date: "2024-02-07"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, echo = FALSE)
```

## Question #1

Suppose x = 1.1, a = 2.2, and b = 3.3. Assign each expression to the value of the variable z and print the value stored in z.
  
$x^{a^b}$  
$(x^a)^b$  
$3x^3+2x^2+1$  

```{r echo=TRUE, results=TRUE}
# Use the assignment operator
x <- 1.1 
a <- 2.2
b <- 3.3

z = x^{a^b}
print(z)

z = (x^a)^b
print(z)

z = 3*x^3+2*x^2+1
print(z)

```

## Question #2

Using the rep and seq functions, create the following vectors:

(1,2,3,4,5,6,7,8,7,6,5,4,3,2,1)  
(1,2,2,3,3,3,4,4,4,4,5,5,5,5,5)  
(5,4,4,3,3,3,2,2,2,2,1,1,1,1,1)  

```{r echo=TRUE, results=TRUE}
z <- c(seq(0,8),seq(7,1))
print(z)

```
```{r echo=TRUE, results=TRUE}
my_vec <- c(1,2,3,4,5)
my_vec <- rep(x=my_vec,times=my_vec)
print(my_vec)
```
```{r echo=TRUE, results=TRUE}
my_vec <- c(5,4,3,2,1)
my_vec <- rep(x=my_vec,times=c(1,2,3,4,5))
print(my_vec)
```

## Question #3


Create a vector of two random uniform numbers. In a spatial map, these can be interpreted as x and y coordinates that give the location of an individual (such as a marked forest tree in a plot that has been mapped). Using one of R’s inverse trigonometry functions `asin()`, `acos()`, or `atan()`, convert these numbers into polar coordinates (If you don’t know what polar coordinates are, read about them on the web here, here, or in your calculus textbook)  

```{r echo=TRUE, results=TRUE}
set.seed(100) #sets the random number 
ranNum <- runif(n=2, min=-90, max=90) #stores random numbers as ranNum
x <- ranNum[1] #assigns first random number as x
y <- ranNum[2] #assigns second random number as y

r <-  sqrt((x*x) + (y*y)) #use conversion formula to calculate r
print(r)
r # Or can just type r and get the same thing
theta <- atan(y/x) #use conversion formula to calculate theta
print(theta) 
theta # Or can just type theta and that does the same thing

```


## Question #4

Create a vector queue <- c("sheep", "fox", "owl", "ant") where queue represents the animals that are lined up to enter Noah’s Ark, with the sheep at the front of the line. Using R expressions, update queue as:

the serpent arrives and gets in line;
the sheep enters the ark;
the donkey arrives and talks his way to the front of the line;
the serpent gets impatient and leaves;
the owl gets bored and leaves;
the aphid arrives and the ant invites him to cut in line.
Finally, determine the position of the aphid in the line.


```{r echo=TRUE, results=TRUE}

Ark <-c("sheep","fox","owl","ant") #sets up my queue
my_Ark <-Ark[c(1,2,3,4)]
print(my_Ark)
my_Ark <- c(my_Ark, "serpent")
my_Ark
my_Ark <- c(my_Ark[-1])
my_Ark
my_Ark[5] <- "donkey"
my_Ark
my_Ark <- my_Ark[c(1,2,3,5,4)]
my_Ark
my_Ark <- my_Ark[c(1,2,4,3,5)]
my_Ark
my_Ark <- my_Ark[c(1,3,2,4,5)]
my_Ark
my_Ark <- my_Ark[c(2,1,3,4,5)]
my_Ark
my_Ark <- c(my_Ark[-5])
my_Ark
my_Ark <- c(my_Ark[-3])
my_Ark
my_Ark <- c(my_Ark, "aphid")
my_Ark
my_Ark <- my_Ark[c(1,2,4,3)]
my_Ark
which(my_Ark == 'aphid')
```

## Question #5

Use R to create a vector of all of the integers from 1 to 100 that are not divisible by 2, 3, or 7.  

```{r echo=TRUE, results=TRUE}
z <- 1:100 # creates a vector from 1-100
my_numbers <- z[!(z %% 2 == 0 | z %% 3 == 0 | z %% 7 == 0)] # checks for numbers that are divisible by 2, 7, and 3 (specify 0 as remainder), the pipe separating each expression means "or" because trying to exclude any number that is not divisible by 2,3, or 7, and then the ! excludes those numbers from the vector
print(my_numbers) # prints the result


```
