---
title: "Homework #4: Atomic Vectors"
author: "Nicole R Gorman"
date: "2024-02-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, echo = FALSE)
```

## Question #1

Suppose x = 1.1, a = 2.2, and b = 3.3. Assign each expression to the value of the variable z and print the value stored in z.
  
$x^{a^b}$  
$(x^a)^b$  
$3x^3+2x^2+1$  

```{r echo=TRUE, results=TRUE}

# Use the assignment operator
x <- 1.1 
a <- 2.2
b <- 3.3

z = x^{a^b}
print(z)

z = (x^a)^b
print(z)

z = 3*x^3+2*x^2+1
print(z)

```

## Question #2

Using the rep and seq functions, create the following vectors:

(1,2,3,4,5,6,7,8,7,6,5,4,3,2,1)  
(1,2,2,3,3,3,4,4,4,4,5,5,5,5,5)  
(5,4,4,3,3,3,2,2,2,2,1,1,1,1,1)  

```{r echo=TRUE, results=TRUE}
z <- c(seq(0,8),seq(7,1))
print(z)

```
```{r echo=TRUE, results=TRUE}
my_vec <- c(1,2,3,4,5)
my_vec <- rep(x=my_vec,times=my_vec)
print(my_vec)
```
```{r echo=TRUE, results=TRUE}
my_vec <- c(5,4,3,2,1)
my_vec <- rep(x=my_vec,times=c(1,2,3,4,5))
print(my_vec)
```

## Question #3


Create a vector of two random uniform numbers. In a spatial map, these can be interpreted as x and y coordinates that give the location of an individual (such as a marked forest tree in a plot that has been mapped). Using one of R’s inverse trigonometry functions `(asin()`, `acos()`, or `atan())`, convert these numbers into polar coordinates (If you don’t know what polar coordinates are, read about them on the web here, here, or in your calculus textbook)  

```{r echo=TRUE, results=TRUE}
runif(n=2, min=-90, max=90)
x <- seq(1:2)
x <- 1
y <- 2
asin(x)
asin(y)
```

## Question #4

Create a vector queue <- c("sheep", "fox", "owl", "ant") where queue represents the animals that are lined up to enter Noah’s Ark, with the sheep at the front of the line. Using R expressions, update queue as:

the serpent arrives and gets in line;
the sheep enters the ark;
the donkey arrives and talks his way to the front of the line;
the serpent gets impatient and leaves;
the owl gets bored and leaves;
the aphid arrives and the ant invites him to cut in line.
Finally, determine the position of the aphid in the line.


```{r echo=TRUE, results=TRUE}


```

## Question #5

Use R to create a vector of all of the integers from 1 to 100 that are not divisible by 2, 3, or 7. You will need one of the arithmetic operators on this cheat sheet.  

```{r echo=TRUE, results=TRUE}


```
