---
title: "Homework 7"
author: "Nicole R. Gorman"
date: "2024-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Creating Fake Data Sets To Explore Hypotheses
* Think about an ongoing study in your lab (or a paper you have read in a different class), and decide on a pattern that you might expect in your experiment if a specific hypothesis were true.

* To start simply, assume that the data in each of your treatment groups follow a normal distribution. Specify the sample sizes, means, and variances for each group that would be reasonable if your hypothesis were true. You may need to consult some previous literature and/or an expert in the field to come up with these numbers.  
<br> 

**Study background**  
*Cells must move materials in order to maintain metabolic processes. Some intracellular movement of materials is accomplished with endosomal trafficking. In plants, defects in trafficking pathways can lead to mutant phenotypes such as short roots. A protein complex known as the retriever has been well characterized in humans, but has yet to be completely explored in plants. Building on previous work in the lab, I am currently working on continuing to the characterize the retreiver in Arabidopsis thaliana. One approach to this question is identifying candidate Arabidopsis mutants with short root phenotypes. Identification of mutant phenotypes provides a foundation for exploring the proteins that may be components of the retreiver in this model species. The data set used in this analysis was generated from the observation of root length phenotypes of putative components of retriever.*

<br>
* Open libraries
```{r open libraries, message=FALSE}
library(ggplot2) # for graphics
library(MASS) # for maximum likelihood estimation
library(dplyr)
```

*This data is from a characterization of a mutant phenotype in Arabidopsis thaliana. The typical sample size for analysis of root length phenotypes is n = 10 plants/genotype/treatment. For this dataset, there are 3 independent biological replicates, but only the first replicate is included with this analysis to help simplify. For this particular dataset, the data need to be reformatted. In the future, data will be formatted differently to simplify downstream alalyses.*  

* Read in and plot the data  

```{r echo=TRUE, warnings=FALSE, results=TRUE}
# Read in the data
z <- read.csv("/Users/nicolegorman/Documents/UVM_Research/UVM Rotation Project/ACC Trials/CBIO_ACCTrials.csv",header=TRUE,sep=",")

# Use regular expressions to subset genotype and then add this column to the dataset
# SEARCH: (\w+_)(\w+),(.*) REPLACE: \2

vecg <- read.csv("/Users/nicolegorman/Documents/UVM_Research/UVM Rotation Project/ACC Trials/CBIO_ACCTrials_Genotypes.csv",header=TRUE,sep=",")

# subset the genotype column
genotype <- vecg$Genotype

# add vecg as a column to the dataset
z$Genotype <- genotype

# get the structure and summary metrics of the new dataset
str(z)
summary(z)

# take a peek to make sure that the genotype column is all there
head (z)
tail (z)

# Remove NAs here
z<-na.omit(z)
# get the structure of z wuithout the NAs
str(z)

# Use myVar for my response variable (length) so I can run the code with any response variable by 
# changing the variable name to myVar
z$myVar <- z$Length

# define variables for easier downstream analysis
ID <- seq_len(nrow(z)) # creates a sequence from 1:n (if n > 0!)
varA <- z$Plant.ID
varB <- z$Genotype
myVar <- z$Length

head(ID)
head (varA)
head(varB)
head (myVar)

# Plot histogram of data
p1 <- ggplot(data=z, aes(x=myVar, y=..density..)) +
  geom_histogram(color="grey60",fill="cornsilk",linewidth=0.2) 
print(p1)

# Add empirical density curve to smooth out the profile of the distribution
p1 <-  p1 +  geom_density(linetype="dotted",size=0.75)
print(p1)

# Get maximum likelihood parameters for the normal distribution
normPars <- fitdistr(z$myVar,"normal")
print(normPars)

str(normPars)
normPars$estimate["mean"] # get a named attribute, I think this might be redundant with below
normPars$estimate["sd"] # get a named attribute, again....I think this might be redundant with below

```

Specify the sample sizes, means, and variances for each group that would be reasonable if your hypothesis were true.  

*I am going to use the parameters from my data set to generate distributions for each variable. Based on this pilot study, I will use the null hypothesis, that genotype has no effect on root growth in mutant plants.*

```{r echo=TRUE, warnings=FALSE, results=TRUE}
# Get maximum likelihood parameters
m_length <- mean(z$myVar)
print(m_length)

sd_length <- sd(z$myVar)
print(sd_length)

#Plot normal probability density
meanML <- normPars$estimate["mean"]
sdML <- normPars$estimate["sd"]

xval <- seq(0,max(z$myVar),len=length(z$myVar))

stat <- stat_function(aes(x = xval, y = ..y..), fun = dnorm, colour="red", n = length(z$myVar), args = list(mean = m_length, sd = sd_length))
p1 + stat

# Assume that each of the treatments follows a normal distribution  
# Get maximum likelihood (ML) parameters for normal distribution for each treatment
# First get the maximum likelihood parameters for a normal distribution fitted to these data by calling fitsdir 
# Fit normal distributions for each group defined by 'Genotype'

# tapply splits the data in z$Length into subsets based on the values in z$Plant.ID
# Then, it applies the fitdistr() to each subset of data to fit a normal distribution
# norm_pars contains a list of parameter estimates for the normal distribution fitted to the lengths of each plant
# Each element of the list corresponds to a unique value of z$Plant.ID
# parameters of the fitted normal distribution for the corresponding subset of root lengths are stored in each element

# Fit normal distributions for each treatment (genotype/hormone treatment)
# tapply applies a function to subsets of a vector or array, based on one or more factors
norm_pars <- tapply(z$Length, z$Plant.ID, function(x) fitdistr(x, "normal"))
norm_pars <- tapply(z$Length, z$Genotype, function(x) fitdistr(x, "normal"))

# these are rough estimates of root length if using both ACC and genotype
# Save this information for next time :)
# Skip these data and see below for parameter estimates if using JUST genotype

# mean         sd    
######### 0nM
# $`1_1_0nM_WT`
# 10.700000    4.472109 
# ( 1.999988) ( 1.414205)
# $`1_2_0nM_ccdc22`
# 11.328000    4.436014 
# ( 1.983846) ( 1.402791)
# $`1_1_0nM_ccdc93`
# 13.874000    4.813488 
# ( 2.152657) ( 1.522158)
# $`1_2_0nM_ccdc22ccdc93`
# 13.704000    5.159452 
# ( 2.307377) ( 1.631562)
# $`1_2_0nM_ccdc22RFP`
# 14.682000    5.267760 
# ( 2.355814) ( 1.665812)
# $`1_1_0nM_ccdc93RFP`
# 14.800200    5.222040 
# ( 2.335368) ( 1.651354)


######### 10nM
# $`1_1_10nM_WT`
# 13.582000    5.310267 
# ( 2.374824) ( 1.679254)
# $`1_1_10nM_ccdc22`
# 7.300000   3.756972 
# (1.680169) (1.188059)
# $`1_1_10nM_ccdc93`
# 9.972000   3.797154 
# (1.698139) (1.200765)
# $`1_1_10nM_ccdc22ccdc93`
# 8.1960000   1.8234100 
# (0.8154537) (0.5766129)
# $`1_1_10nM_ccdc22RFP`
# 12.638000    4.624255 
# ( 2.068030) ( 1.462318)
# $`1_1_10nM_ccdc93RFP`
# 10.2852000    2.2828781 
# ( 1.0209341) ( 0.7219095)

######### 100nM
# $`1_1_100nM_WT`
# 6.7160000   1.9861279 
# (0.8882234) (0.6280688)
# $`1_1_100nM_ccdc22`
# 7.6620000   3.1125706 
# (1.3919839) (0.9842813)
# $`1_1_100nM_ccdc93`
# 9.5980000   3.1054880 
# (1.3888165) (0.9820415)
# $`1_1_100nM_ccdc22ccdc93`
# 9.934000   3.391782 
# (1.516851) (1.072576)
# $`1_1_100nM_ccdc22RFP`
# 10.2300000    2.3405384 
# ( 1.0467206) ( 0.7401432)
# $`1_1_100nM_ccdc93RFP`
# 7.1680000   1.8515218 
# (0.8280257) (0.5855026)

######### 1000nM (1uM)
# $`1_1_1uM_WT`
# 10.0180000    1.8260164 
# ( 0.8166194) ( 0.5774371)
# $`1_1_1uM_ccdc22`
# 8.3880000   1.7285069 
# (0.7730118) (0.5466019)
# $`1_1_1uM_ccdc93`
# 10.4520000    2.5107003 
# ( 1.1228193) ( 0.7939531)
# $`1_1_1uM_ccdc22ccdc93`
# 8.1420000   2.2386013 
# (1.0011330) (0.7079079)
# $`1_1_1uM_ccdc22RFP`
# 9.9660000   1.9445884 
# (0.8696464) (0.6149328)
# $`1_1_1uM_ccdc93RFP`
# 9.6238000   2.0015200 
# (0.8951069) (0.6329362)

# View the ML parameters for each treatment
print(norm_pars)
summary(norm_pars)
# str (norm_pars)...woah!

# These are the ML parameters for genotype only

# $WT
# mean          sd    
# 11.8440500    4.6060560 
# ( 0.3256973) ( 0.2303028)
#
# $ccdc22
# mean         sd    
# 9.5992222   4.8467498 
# (0.3612554) (0.2554461)
# 
# $ccdc93
# mean          sd    
# 11.3043333    4.7605129 
# ( 0.3548277) ( 0.2509011)
#
# $ccdc22ccdc93
# mean          sd    
# 10.0921805    4.0788841 
# ( 0.3536841) ( 0.2500924)
# 
# $ccdc22RFP
# mean          sd    
# 10.9336154    4.2258974 
# ( 0.3706357) ( 0.2620790)
# 
# $ccdc93RFP
# mean          sd    
# 10.9235856    4.2714427 
# ( 0.3174938) ( 0.2245020)

# the data is not set up for this, but it is actually the correct way to analyze this full dataset
# Come back and do the analysis another time after editing dataset
# Need to make treatment its own column and add all three reps
```

# Generate Random Data Set
Using the methods we have covered in class, write code to create a random data set that has these attributes. 

*For this analysis, I simplified and fit normal distribution for each genotype only, ignoring the hormone treatment.*

```{r echo=TRUE, warnings=FALSE, results=TRUE}

# this set of calls creates individual normal distributions for each variable
rWT <- rnorm(50,11.8,4.6)
mean(rWT)  
sd(rWT)  

r22 <- rnorm(50,9.6,4.8)
mean(r22)  
sd(r22)  

r93 <- rnorm(50,11.3,4.8)
mean(r93)  
sd(r93)  

r2293 <- rnorm(50,10.1,4.1)
mean(r2293)  
sd(r2293)  

r22R <- rnorm(50,10.9,4.2)
mean(r22R)
sd(r22R)  

r93R <- rnorm(50,10.9,4.3)
mean(r93R)  
sd(r93R) 

# string these into a single vector

resVar <- c(rWT,r22,r93,r2293,r22R,r93R)
#head (resVar)

# associate each response variable with the appropriate genotype, n=50
rGen <- c(rep("WT", 50), rep("CCDC22", 50), rep("CCDC93", 50), 
          rep("CCDC22CCDC93", 50), rep("CCDC22RFP", 50), 
          rep("CCDC93RFP", 50))

head(rGen)
tail(rGen)

# this is another way to do the same thing
# rep(nName, each=50)

```

# Create Data Frame
Organize these data into a data frame with the appropriate structure.  

*BUT...now that I have written the code above, I don't think I am actually going to use it. I will create the dataframe and generate the random numbers using the same call from class.*

```{r echo=TRUE, warnings=FALSE, results=TRUE}
# Data frame construction for one-way ANOVA
# create a dataframe of random variables

nGroup <- 6 # number of treatment groups
nName <- c("WT","CCDC22", "CCDC93", "CCDC22CCDC93", "CCDC22RFP", "CCDC93RFP") # names of groups
nSize <- c(50,50,50,50,50,50) # number of observations in each group, these are all equal in the simulated dataset
# the mean and variance are estimated using pilot trial numbers

nMean <- c(11.8,9.6,11.3,10.1,10.9,10.9) # mean of each group
nSD <- c(4.6,4.8,4.8,4.1,4.2,4.3) # standard deviation of each group

# id vector for each row, just numbers each row for organization and/or reference
ID <- 1:(sum(nSize)) 

# create a vector by concatenating six sets of random normal variables
# reVar is the reponse variable, in this case it is root length
resVar <- c(rnorm(n=nSize[1],mean=nMean[1],sd=nSD[1]), 
            rnorm(n=nSize[2],mean=nMean[2],sd=nSD[2]),
            rnorm(n=nSize[3],mean=nMean[3],sd=nSD[3]), 
            rnorm(n=nSize[4],mean=nMean[4],sd=nSD[4]),
            rnorm(n=nSize[5],mean=nMean[5],sd=nSD[5]),
            rnorm(n=nSize[6],mean=nMean[6],sd=nSD[6]))
# sets up single column with all data, this is now in long form

# create dataframe for downstream ANOVA
TGroup <- rep(nName,nSize)
ANOdata <- data.frame(ID,TGroup,resVar) 
str(ANOdata)
print(ANOdata) # prints the dataset

```

# Run ANOVA
Now write code to analyze the data...probably as an ANOVA or regression analysis, but possibly as a logistic regression or contingency table analysis. 

```{r echo=TRUE, warnings=FALSE, results=TRUE}

# Run basic ANOVA
ANOmodel <- aov(resVar~TGroup,data=ANOdata)
print(ANOmodel)
print(summary(ANOmodel))
z <- summary(ANOmodel)
str(z)
aggregate(resVar~TGroup,data=ANOdata,FUN=mean)
unlist(z)
unlist(z)[7] # F value1 3.460314 
ANOsum <- list(Fval=unlist(z)[7],probF=unlist(z)[9])
ANOsum  # Pr(>F)1     0.004699716 
```

# Create a Box Plot
Write code to generate a useful graph of the data.

```{r echo=TRUE, warnings=FALSE, results=TRUE}

# Use ggplot to visualize the ANOVA data
ANOPlot <- ggplot(data=ANOdata) + 
  aes(x=TGroup,y=resVar,fill=TGroup) +
  geom_boxplot()
print(ANOPlot)

```
<br>
# Run the Simulation
Try running your analysis multiple times to get a feeling for how variable the results are with the same parameters, but different sets of random numbers.

```{r echo=TRUE, warnings=FALSE, results=TRUE}
nGroup <- 6 # number of treatment groups
nName <- c("WT","CCDC22", "CCDC93", "CCDC22CCDC93", "CCDC22RFP", "CCDC93RFP") # names of groups
nSize <- c(50,50,50,50,50,50) # number of observations in each group, these are all equal in the simulated dataset
# the mean and variance are estimated using pilot trial numbers

nMean <- c(11.8,9.6,11.3,10.1,10.9,10.9) # mean of each group
nSD <- c(4.6,4.8,4.8,4.1,4.2,4.3) # standard deviation of each group

# id vector for each row, just numbers each row for organization and/or reference
ID <- 1:(sum(nSize)) 

# create a vector by concatenating six sets of random normal variables
# reVar is the response variable, root length
resVar <- c(rnorm(n=nSize[1],mean=nMean[1],sd=nSD[1]), 
            rnorm(n=nSize[2],mean=nMean[2],sd=nSD[2]),
            rnorm(n=nSize[3],mean=nMean[3],sd=nSD[3]), 
            rnorm(n=nSize[4],mean=nMean[4],sd=nSD[4]),
            rnorm(n=nSize[5],mean=nMean[5],sd=nSD[5]),
            rnorm(n=nSize[6],mean=nMean[6],sd=nSD[6]))
# sets up single column with all data, this is now in long form

TGroup <- rep(nName,nSize) # TGruop = genotype
ANOdata <- data.frame(ID,TGroup,resVar) # create dataframe
str(ANOdata)
# print(ANOdata)

# Run basic ANOVA
# the aggregate function calculates the mean of resVar (root length)for each level 
# of TGroup (genotype) in the ANOdata data frame
ANOmodel <- aov(resVar~TGroup,data=ANOdata)
print(ANOmodel)
print(summary(ANOmodel)) # Estimated effects may be unbalanced, not sure if this message is a problem?
z <- summary(ANOmodel)
str(z)
aggregate(resVar~TGroup,data=ANOdata,FUN=mean)
unlist(z)
unlist(z)[7] # F value1 multiple run results 3.460314 |1.790044  | 3.486434 | 2.642683 | 1.430171 | 2.554768 
ANOsum <- list(Fval=unlist(z)[7],probF=unlist(z)[9])
ANOsum  # Pr(>F)1 multiple run results 0.004699716 | 0.1147189 | 0.004460621 | 0.02346638 | 0.2132236 | 0.02778633 
# Pr(>F) p-value associated with the F-statistic 
# If Pr(>F) is < 0.05, then at least one of the groups is different 

# Use ggplot to visualize the ANOVA data
ANOPlot <- ggplot(data=ANOdata) + 
  aes(x=TGroup,y=resVar,fill=TGroup) +
  geom_boxplot()
print(ANOPlot)

```

Commenting out this first method of generating random normal numbers for each genotype separately 

```{r echo=TRUE, warnings=FALSE, results=TRUE}

# this set of calls creates individual normal distributions for each variable
# rWT <- rnorm(50,11.8,4.6)
# mean(rWT)  
# sd(rWT)  
# 
# r22 <- rnorm(50,9.6,4.8)
# mean(r22)  
# sd(r22)  
# 
# r93 <- rnorm(50,11.3,4.8)
# mean(r93)  
# sd(r93)  
# 
# r2293 <- rnorm(50,10.1,4.1)
# mean(r2293)  
# sd(r2293)  
# 
# r22R <- rnorm(50,10.9,4.2)
# mean(r22R)
# sd(r22R)  
# 
# r93R <- rnorm(50,10.9,4.3)
# mean(r93R)  
# sd(r93R) 
# 
# # string these into a single vector
# 
# resVar <- c(rWT,r22,r93,r2293,r22R,r93R)
# #head (resVar)
# 
# # associate each response variable with the appropriate genotype, n=50
# rGen <- c(rep("WT", 50), rep("CCDC22", 50), rep("CCDC93", 50), 
#           rep("CCDC22CCDC93", 50), rep("CCDC22RFP", 50), 
#           rep("CCDC93RFP", 50))
# 
```

# Adjust the Mean and sd to Explore the Experimental Design
Now begin adjusting the means of the different groups. 
*First, the mean was adjusted to fit the model. Then, the code was run with smaller root length differences to determine how much of a difference is statically different. Finally, the sd was adjusted to an average across all genotypes to determine the effect of variation with each genotype on the outcome.* 

```{r echo=TRUE, warnings=FALSE, results=TRUE}

# Adjusted Parameters #1: Adjust the mean to fit the expected model.

# Data frame construction for one-way ANOVA
# create a dataframe of random variables

nGroup <- 6 # number of treatment groups
nName <- c("WT","CCDC22", "CCDC93", "CCDC22CCDC93", "CCDC22RFP", "CCDC93RFP") # names of groups
nSize <- c(50,50,50,50,50,50) # number of observations in each group, these are all equal in the simulated dataset
# the mean and variance are estimated using pilot trial numbers

nMean <- c(16,8,8,8,16,16) # adjust the mean of each group
nSD <- c(4.6,4.8,4.8,4.1,4.2,4.3) # leave the standard deviation of each group as estimated by the pilot

# id vector for each row, just numbers each row for organization and/or reference
ID <- 1:(sum(nSize)) 

# create a vector by concatenating six sets of random normal variables
resVar <- c(rnorm(n=nSize[1],mean=nMean[1],sd=nSD[1]), 
            rnorm(n=nSize[2],mean=nMean[2],sd=nSD[2]),
            rnorm(n=nSize[3],mean=nMean[3],sd=nSD[3]), 
            rnorm(n=nSize[4],mean=nMean[4],sd=nSD[4]),
            rnorm(n=nSize[5],mean=nMean[5],sd=nSD[5]),
            rnorm(n=nSize[6],mean=nMean[6],sd=nSD[6]))
# sets up single column with all data, this is now in long form

# create dataframe
TGroup <- rep(nName,nSize)
ANOdata <- data.frame(ID,TGroup,resVar) 
str(ANOdata)
# print(ANOdata)

# Run basic ANOVA
ANOmodel <- aov(resVar~TGroup,data=ANOdata)
print(ANOmodel)
print(summary(ANOmodel))
z <- summary(ANOmodel)
str(z)
aggregate(resVar~TGroup,data=ANOdata,FUN=mean)
unlist(z)
unlist(z)[7] # F value1 3.460314 
ANOsum <- list(Fval=unlist(z)[7],probF=unlist(z)[9])
ANOsum  # Pr(>F)1     0.004699716 
# Pr(>F) p-value associated with the F-statistic 
# If Pr(>F) is < 0.05, then at least one of the groups is different 

# Use ggplot to visualize the ANOVA data
ANOPlot <- ggplot(data=ANOdata) + 
  aes(x=TGroup,y=resVar,fill=TGroup) +
  geom_boxplot()
print(ANOPlot)

# Adjusted Parameters #2: Does a smaller (2mm) mean root length difference still generate significant values?

nGroup <- 6 # number of treatment groups
nName <- c("WT","CCDC22", "CCDC93", "CCDC22CCDC93", "CCDC22RFP", "CCDC93RFP") # names of groups
nSize <- c(50,50,50,50,50,50) # number of observations in each group, these are all equal in the simulated dataset
# the mean and variance are estimated using pilot trial numbers

nMean <- c(10,8,8,8,10,10) # adjust the mean of each group
nSD <- c(4.6,4.8,4.8,4.1,4.2,4.3) # leave the standard deviation of each group as estimated by the pilot

# id vector for each row, just numbers each row for organization and/or reference
ID <- 1:(sum(nSize)) 

# create a vector by concatenating six sets of random normal variables
resVar <- c(rnorm(n=nSize[1],mean=nMean[1],sd=nSD[1]), 
            rnorm(n=nSize[2],mean=nMean[2],sd=nSD[2]),
            rnorm(n=nSize[3],mean=nMean[3],sd=nSD[3]), 
            rnorm(n=nSize[4],mean=nMean[4],sd=nSD[4]),
            rnorm(n=nSize[5],mean=nMean[5],sd=nSD[5]),
            rnorm(n=nSize[6],mean=nMean[6],sd=nSD[6]))
# sets up single column with all data, this is now in long form

# create dataframe
TGroup <- rep(nName,nSize)
ANOdata <- data.frame(ID,TGroup,resVar) 
str(ANOdata)
# print(ANOdata)

# Run basic ANOVA
ANOmodel <- aov(resVar~TGroup,data=ANOdata)
print(ANOmodel)
print(summary(ANOmodel))
z <- summary(ANOmodel)
str(z)
aggregate(resVar~TGroup,data=ANOdata,FUN=mean)
unlist(z)
unlist(z)[7] # F value1 of multiple runs  3.460314 | 3.057447 | 4.37688 
ANOsum <- list(Fval=unlist(z)[7],probF=unlist(z)[9])
ANOsum  # Pr(>F)1 of multiple runs 0.004699716 | 0.01045411 | 0.0007376741
# Pr(>F) p-value associated with the F-statistic 
# If Pr(>F) is < 0.05, then at least one of the groups is different 

# Use ggplot to visualize the ANOVA data
ANOPlot <- ggplot(data=ANOdata) + 
  aes(x=TGroup,y=resVar,fill=TGroup) +
  geom_boxplot()
print(ANOPlot)

# Adjusted Parameters #3: Does an even smaller difference (1mm) in mean root length still generate significant values?

nGroup <- 6 # number of treatment groups
nName <- c("WT","CCDC22", "CCDC93", "CCDC22CCDC93", "CCDC22RFP", "CCDC93RFP") # names of groups
nSize <- c(50,50,50,50,50,50) # number of observations in each group, these are all equal in the simulated dataset
# the mean and variance are estimated using pilot trial numbers

nMean <- c(10,9,9,9,10,10) # adjust the mean of each group
nSD <- c(4.6,4.8,4.8,4.1,4.2,4.3) # leave the standard deviation of each group as estimated by the pilot

# id vector for each row, just numbers each row for organization and/or reference
ID <- 1:(sum(nSize)) 

# create a vector by concatenating six sets of random normal variables
resVar <- c(rnorm(n=nSize[1],mean=nMean[1],sd=nSD[1]), 
            rnorm(n=nSize[2],mean=nMean[2],sd=nSD[2]),
            rnorm(n=nSize[3],mean=nMean[3],sd=nSD[3]), 
            rnorm(n=nSize[4],mean=nMean[4],sd=nSD[4]),
            rnorm(n=nSize[5],mean=nMean[5],sd=nSD[5]),
            rnorm(n=nSize[6],mean=nMean[6],sd=nSD[6]))
# sets up single column with all data, this is now in long form

# create dataframe
TGroup <- rep(nName,nSize)
ANOdata <- data.frame(ID,TGroup,resVar) 
str(ANOdata)
# print(ANOdata)

# Run basic ANOVA
ANOmodel <- aov(resVar~TGroup,data=ANOdata)
print(ANOmodel)
print(summary(ANOmodel))
z <- summary(ANOmodel)
str(z)
aggregate(resVar~TGroup,data=ANOdata,FUN=mean)
unlist(z)
unlist(z)[7] # F value1 of multiple runs  0.8566817 | 2.737496 | 2.725934
ANOsum <- list(Fval=unlist(z)[7],probF=unlist(z)[9])
ANOsum  # Pr(>F)1 of multiple runs 0.5106597 | 0.01953694 | 0.01997955
# Pr(>F) p-value associated with the F-statistic 
# If Pr(>F) is < 0.05, then at least one of the groups is different 

# Use ggplot to visualize the ANOVA data
ANOPlot <- ggplot(data=ANOdata) + 
  aes(x=TGroup,y=resVar,fill=TGroup) +
  geom_boxplot()
print(ANOPlot)

# Adjusted Parameters #4: Set the variation within genotypes as a constant.

# Data frame construction for one-way ANOVA
# create a dataframe of random variables

nGroup <- 6 # number of treatment groups
nName <- c("WT","CCDC22", "CCDC93", "CCDC22CCDC93", "CCDC22RFP", "CCDC93RFP") # names of groups
nSize <- c(50,50,50,50,50,50) # number of observations in each group, these are all equal in the simulated dataset
# the mean and variance are estimated using pilot trial numbers

nMean <- c(11.8,9.6,11.3,10.1,10.9,10.9) # original means
nSD <- c(4.5,4.5,4.5,4.5,4.5,4.5) # set the standard deviation as a constant across all genotypes

# id vector for each row, just numbers each row for organization and/or reference
ID <- 1:(sum(nSize)) 

# create a vector by concatenating six sets of random normal variables
resVar <- c(rnorm(n=nSize[1],mean=nMean[1],sd=nSD[1]), 
            rnorm(n=nSize[2],mean=nMean[2],sd=nSD[2]),
            rnorm(n=nSize[3],mean=nMean[3],sd=nSD[3]), 
            rnorm(n=nSize[4],mean=nMean[4],sd=nSD[4]),
            rnorm(n=nSize[5],mean=nMean[5],sd=nSD[5]),
            rnorm(n=nSize[6],mean=nMean[6],sd=nSD[6]))
# sets up single column with all data, this is now in long form

# create dataframe
TGroup <- rep(nName,nSize)
ANOdata <- data.frame(ID,TGroup,resVar) 
str(ANOdata)
# print(ANOdata)

# Run basic ANOVA
ANOmodel <- aov(resVar~TGroup,data=ANOdata)
print(ANOmodel)
print(summary(ANOmodel))
z <- summary(ANOmodel)
str(z)
aggregate(resVar~TGroup,data=ANOdata,FUN=mean)
unlist(z)
unlist(z)[7] # F value1 of multiple runs  1.183123 | 4.613077 
ANOsum <- list(Fval=unlist(z)[7],probF=unlist(z)[9])
ANOsum  # Pr(>F)1 of multiple runs  0.3174042 | 0.0004555719 
# Pr(>F) p-value associated with the F-statistic 
# If Pr(>F) is < 0.05, then at least one of the groups is different 
# adjusting the sd seems to result in the data more more often failing to reject the null

# Use ggplot to visualize the ANOVA data
ANOPlot <- ggplot(data=ANOdata) + 
  aes(x=TGroup,y=resVar,fill=TGroup) +
  geom_boxplot()
print(ANOPlot)

```

Given the sample sizes you have chosen, how small can the differences between the groups be (the “effect size”) for you to still detect a significant pattern (p < 0.05)?

*My understanding is that it is valuable to look at both the F-value and the p-value. The F-value is used to test the null hypothesis that the means of several groups are equal. In this case that means that all genotypes have the same root length. If the F-value is large and the associated p-value is small (P < 0.05), it suggests that there is a significant difference between the means of at least two groups.*

*A difference of 2mm in average root length among the genotypes reliably generated a significant difference. However, a difference of 1mm  generated a significant result only some of the time. This implies that there is enough variability in average root length within each genotype that some value between 1-2 mm average root length difference is required to reject the null. In other words, a difference of greater than 1mm is needed to conclude that there is a difference in root length between wildtype and the mutant genotypes.*

# Explore Sample Size

Alternatively, for the effect sizes you originally hypothesized, what is the minimum sample size you would need in order to detect a statistically significant effect? Again, run the model a few times with the same parameter set to get a feeling for the effect of random variation in the data.

*A sample size of 75 is needed to generate a reliable statistically significant results with the give ML parameters.This is a large sample size compared to the typical design that has been used in the past. This experiment needs an additional trails and perhaps an adjustment of the treatments to detemine whether root length is a mutant phenotype for CCDC93 and CCDC22. For example, seedlings may need to be grown with an opsmotic stress treatment to use the root length phenotype for analysis.*


```{r echo=TRUE, warnings=FALSE, results=TRUE}

################################
# Adjust sample size to n=10
################################

nGroup <- 6 # number of treatment groups
nName <- c("WT","CCDC22", "CCDC93", "CCDC22CCDC93", "CCDC22RFP", "CCDC93RFP") # names of groups
nSize <- c(10,10,10,10,10,10) # number of observations in each group, these are all equal in the simulated dataset
# the mean and variance are estimated using pilot trial numbers

nMean <- c(11.8,9.6,11.3,10.1,10.9,10.9) # mean of each group
nSD <- c(4.6,4.8,4.8,4.1,4.2,4.3) # standard deviation of each group

# id vector for each row, just numbers each row for organization and/or reference
ID <- 1:(sum(nSize)) 

# create a vector by concatenating six sets of random normal variables
# reVar is the response variable, root length
resVar <- c(rnorm(n=nSize[1],mean=nMean[1],sd=nSD[1]), 
            rnorm(n=nSize[2],mean=nMean[2],sd=nSD[2]),
            rnorm(n=nSize[3],mean=nMean[3],sd=nSD[3]), 
            rnorm(n=nSize[4],mean=nMean[4],sd=nSD[4]),
            rnorm(n=nSize[5],mean=nMean[5],sd=nSD[5]),
            rnorm(n=nSize[6],mean=nMean[6],sd=nSD[6]))
# sets up single column with all data, this is now in long form

TGroup <- rep(nName,nSize) # TGruop = genotype
ANOdata <- data.frame(ID,TGroup,resVar) # create dataframe
str(ANOdata)
# print(ANOdata)

# Run basic ANOVA
# the aggregate function calculates the mean of resVar (root length)for each level 
# of TGroup (genotype) in the ANOdata data frame
ANOmodel <- aov(resVar~TGroup,data=ANOdata)
print(ANOmodel)
print(summary(ANOmodel)) # Estimated effects may be unbalanced, not sure if this message is a problem?
z <- summary(ANOmodel)
str(z)
aggregate(resVar~TGroup,data=ANOdata,FUN=mean)
unlist(z)
unlist(z)[7]  
ANOsum <- list(Fval=unlist(z)[7],probF=unlist(z)[9])
ANOsum  
# Pr(>F) p-value associated with the F-statistic 
# If Pr(>F) is < 0.05, then at least one of the groups is different 
# Multiple run results
#  F value1 
# 0.3651354 | 0.7483566 | 1.354049 
#   Pr(>F)1 
# 0.8701885 | 0.5908783 | 0.2561594 


# Use ggplot to visualize the ANOVA data
ANOPlot <- ggplot(data=ANOdata) + 
  aes(x=TGroup,y=resVar,fill=TGroup) +
  geom_boxplot()
print(ANOPlot)

################################
# Adjust sample size to n=20
################################

nGroup <- 6 # number of treatment groups
nName <- c("WT","CCDC22", "CCDC93", "CCDC22CCDC93", "CCDC22RFP", "CCDC93RFP") # names of groups
nSize <- c(20,20,20,20,20,20) # number of observations in each group, these are all equal in the simulated dataset
# the mean and variance are estimated using pilot trial numbers

nMean <- c(11.8,9.6,11.3,10.1,10.9,10.9) # mean of each group
nSD <- c(4.6,4.8,4.8,4.1,4.2,4.3) # standard deviation of each group

# id vector for each row, just numbers each row for organization and/or reference
ID <- 1:(sum(nSize)) 

# create a vector by concatenating six sets of random normal variables
# reVar is the response variable, root length
resVar <- c(rnorm(n=nSize[1],mean=nMean[1],sd=nSD[1]), 
            rnorm(n=nSize[2],mean=nMean[2],sd=nSD[2]),
            rnorm(n=nSize[3],mean=nMean[3],sd=nSD[3]), 
            rnorm(n=nSize[4],mean=nMean[4],sd=nSD[4]),
            rnorm(n=nSize[5],mean=nMean[5],sd=nSD[5]),
            rnorm(n=nSize[6],mean=nMean[6],sd=nSD[6]))
# sets up single column with all data, this is now in long form

TGroup <- rep(nName,nSize) # TGruop = genotype
ANOdata <- data.frame(ID,TGroup,resVar) # create dataframe
str(ANOdata)
# print(ANOdata)

# Run basic ANOVA
# the aggregate function calculates the mean of resVar (root length)for each level 
# of TGroup (genotype) in the ANOdata data frame
ANOmodel <- aov(resVar~TGroup,data=ANOdata)
print(ANOmodel)
print(summary(ANOmodel)) # Estimated effects may be unbalanced, not sure if this message is a problem?
z <- summary(ANOmodel)
str(z)
aggregate(resVar~TGroup,data=ANOdata,FUN=mean)
unlist(z)
unlist(z)[7] 
ANOsum <- list(Fval=unlist(z)[7],probF=unlist(z)[9])
ANOsum  
# Pr(>F) p-value associated with the F-statistic 
# If Pr(>F) is < 0.05, then at least one of the groups is different 
# Multiple run results
#  F value1 
# 4.254324 | 2.351236  | 1.051566  
#   Pr(>F)1 
# 0.001388849  | 0.04514597 | 0.3911177  

# Use ggplot to visualize the ANOVA data
ANOPlot <- ggplot(data=ANOdata) + 
  aes(x=TGroup,y=resVar,fill=TGroup) +
  geom_boxplot()
print(ANOPlot)

################################
# Adjust sample size to n=75  
################################

nGroup <- 6 # number of treatment groups
nName <- c("WT","CCDC22", "CCDC93", "CCDC22CCDC93", "CCDC22RFP", "CCDC93RFP") # names of groups
nSize <- c(75,75,75,75,75,75) # number of observations in each group, these are all equal in the simulated dataset
# the mean and variance are estimated using pilot trial numbers

nMean <- c(11.8,9.6,11.3,10.1,10.9,10.9) # mean of each group
nSD <- c(4.6,4.8,4.8,4.1,4.2,4.3) # standard deviation of each group

# id vector for each row, just numbers each row for organization and/or reference
ID <- 1:(sum(nSize)) 

# create a vector by concatenating six sets of random normal variables
# reVar is the response variable, root length
resVar <- c(rnorm(n=nSize[1],mean=nMean[1],sd=nSD[1]), 
            rnorm(n=nSize[2],mean=nMean[2],sd=nSD[2]),
            rnorm(n=nSize[3],mean=nMean[3],sd=nSD[3]), 
            rnorm(n=nSize[4],mean=nMean[4],sd=nSD[4]),
            rnorm(n=nSize[5],mean=nMean[5],sd=nSD[5]),
            rnorm(n=nSize[6],mean=nMean[6],sd=nSD[6]))
# sets up single column with all data, this is now in long form

TGroup <- rep(nName,nSize) # TGruop = genotype
ANOdata <- data.frame(ID,TGroup,resVar) # create dataframe
str(ANOdata)
# print(ANOdata)

# Run basic ANOVA
# the aggregate function calculates the mean of resVar (root length)for each level 
# of TGroup (genotype) in the ANOdata data frame
ANOmodel <- aov(resVar~TGroup,data=ANOdata)
print(ANOmodel)
print(summary(ANOmodel)) # Estimated effects may be unbalanced, not sure if this message is a problem?
z <- summary(ANOmodel)
str(z)
aggregate(resVar~TGroup,data=ANOdata,FUN=mean)
unlist(z)
unlist(z)[7] # F value1 
ANOsum <- list(Fval=unlist(z)[7],probF=unlist(z)[9])
ANOsum  # Pr(>F)1 multiple run results
# Pr(>F) p-value associated with the F-statistic 
# If Pr(>F) is < 0.05, then at least one of the groups is different 
# Results from multiple runs
# F value1 
# 3.416318 | 4.355224 | 8.987862
#     Pr(>F)1 
# 0.004859086 | 0.0007015046 | 3.780543e-08

# Use ggplot to visualize the ANOVA data
ANOPlot <- ggplot(data=ANOdata) + 
  aes(x=TGroup,y=resVar,fill=TGroup) +
  geom_boxplot()
print(ANOPlot)

################################
# Adjust sample size to n=100
################################

nGroup <- 6 # number of treatment groups
nName <- c("WT","CCDC22", "CCDC93", "CCDC22CCDC93", "CCDC22RFP", "CCDC93RFP") # names of groups
nSize <- c(100,100,100,100,100,100) # number of observations in each group, these are all equal in the simulated dataset
# the mean and variance are estimated using pilot trial numbers

nMean <- c(11.8,9.6,11.3,10.1,10.9,10.9) # mean of each group
nSD <- c(4.6,4.8,4.8,4.1,4.2,4.3) # standard deviation of each group

# id vector for each row, just numbers each row for organization and/or reference
ID <- 1:(sum(nSize)) 

# create a vector by concatenating six sets of random normal variables
# reVar is the response variable, root length
resVar <- c(rnorm(n=nSize[1],mean=nMean[1],sd=nSD[1]), 
            rnorm(n=nSize[2],mean=nMean[2],sd=nSD[2]),
            rnorm(n=nSize[3],mean=nMean[3],sd=nSD[3]), 
            rnorm(n=nSize[4],mean=nMean[4],sd=nSD[4]),
            rnorm(n=nSize[5],mean=nMean[5],sd=nSD[5]),
            rnorm(n=nSize[6],mean=nMean[6],sd=nSD[6]))
# sets up single column with all data, this is now in long form

TGroup <- rep(nName,nSize) # TGruop = genotype
ANOdata <- data.frame(ID,TGroup,resVar) # create dataframe
str(ANOdata)
# print(ANOdata)

# Run basic ANOVA
# the aggregate function calculates the mean of resVar (root length)for each level 
# of TGroup (genotype) in the ANOdata data frame
ANOmodel <- aov(resVar~TGroup,data=ANOdata)
print(ANOmodel)
print(summary(ANOmodel)) # Estimated effects may be unbalanced, not sure if this message is a problem?
z <- summary(ANOmodel)
str(z)
aggregate(resVar~TGroup,data=ANOdata,FUN=mean)
unlist(z)
unlist(z)[7] # F value1 
ANOsum <- list(Fval=unlist(z)[7],probF=unlist(z)[9])
ANOsum  # Pr(>F)1 multiple run results
# Pr(>F) p-value associated with the F-statistic 
# If Pr(>F) is < 0.05, then at least one of the groups is different 
# Multiple run results
#  F value1 
# 3.913523  | 5.355109   | 4.928941 
#  Pr(>F)1 
# 0.001689342  | 7.961116e-05 | 0.0001980238   

# Use ggplot to visualize the ANOVA data
ANOPlot <- ggplot(data=ANOdata) + 
  aes(x=TGroup,y=resVar,fill=TGroup) +
  geom_boxplot()
print(ANOPlot)
```

If you have time, try repeating this exercise with one of the more sophisticated distributions, such as the gamma or negative binomial (depending on the kind of data you have). You will have to spend some time figuring out by trial and error the parameter values you will need to generate appropriate means and variances of the different groups.

*I'll try at some point, but in the interest of time, I am submitting this now :)*