---
title: "Homework 8"
author: "Nicole R. Gorman"
date: "2024-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data manipulations using the dplyr package


* Open libraries
```{r open libraries, message=FALSE}
library(dplyr)
library(tidyverse)
```

**#1. Examine the structure of the iris data set.**  

```{r echo=TRUE, warnings=FALSE, results=TRUE}
 
data("iris")
str_iris<-str(iris)
#print(str_iris) # does not work
#tried print() so that answer would print to the screen, does not work
```

**How many observations and variables are in the data set?**
```
#'data.frame':	150 obs. of  5 variables:
```

**#2. Create a new data frame iris1 that contains only the species virginica and versicolor with sepal lengths longer than 6 cm and sepal widths longer than 2.5 cm.**  

* Use filter(  ) : pick/subset observations by their values  
* filter automatically excludes NAs, have to ask for them specifically  

```{r echo=TRUE, warnings=FALSE, results=TRUE}

iris1<-filter(iris, Species == "virginica" | Species == "versicolor", "Sepal.Length" > 6, "Sepal.Width" > 2.5)
```
**How many observations and variables are in the data set?**  
```
'data.frame':    100 obs. of  5 variables
```

```{r echo=TRUE, warnings=FALSE, results=TRUE}
str(iris1)
view(iris1)
```
**#3. Now, create a iris2 data frame from iris1 that contains only the columns for Species, Sepal.Length, and Sepal.Width.**  

* Dplyr method-uses select() function, which can use either column # or name

```{r echo=TRUE, warnings=FALSE, results=TRUE}
iris2<-select(iris1, "Species", "Sepal.Length", "Sepal.Width")
```

**How many observations and variables are in the data set?**  
```{r echo=TRUE, warnings=FALSE, results=TRUE}
# get the dataframe structure
str(iris2)
# 'data.frame':	100 obs. of  3 variables
```

**#4. Create an iris3 data frame from iris2 that orders the observations from largest to smallest sepal length.**

```{r echo=TRUE, warnings=FALSE, results=TRUE}
# Create iris3 data frame with ordered observations
iris3 <- iris2 %>% 
arrange(desc(Sepal.Length))
```

**Show the first 6 rows of this data set.**  

```{r echo=TRUE, warnings=FALSE, results=TRUE}
# print(iris3) 
# Use print see the entire data frame

head (iris3)
# Use head to see just the first six rows
```


**#5. Create an iris4 data frame from iris3 that creates a column with a sepal area (length * width) value for each observation.**  
 * use mutate to change dataframe  
 * add new variable by multiplying specified columns  
```{r echo=TRUE, warnings=FALSE, results=TRUE}
iris4 <- iris3 %>% 
mutate(Sepal.Area = Sepal.Length * Sepal.Width)
```

**How many observations and variables are in the data set?**

```{r echo=TRUE, warnings=FALSE, results=TRUE}
str(iris4) # 'data.frame':	100 obs. of  4 variables
# there should be four variables because of the addition of Sepal.Area
# calling structure returns that there are still 100 observations, but now there are 4 variables
```

**#6. Create iris5 that calculates the average sepal length, the average sepal width,and the sample size of the entire iris4 data frame and print iris5.**

```{r echo=TRUE, warnings=FALSE, results=TRUE}
iris5 <- iris4 %>% 
summarise(Avg_Sepal.Length = mean(Sepal.Length),
          Avg_Sepal.Width = mean(Sepal.Width),
          SampleSize = n())
          print(iris5)
# use summarise() to create a new data frame with summary statistics
# the output is a single row summarizing all observations in the input
# it will contain one column for each of the specified summary statistics

# Avg_Sepal.Length Avg_Sepal.Width SampleSize
#     1            6.262           2.872        100
# makes sense that there are still 100 observations, should not change
```          

**#7. Finally, create iris6 that calculates the average sepal length, the average sepal width, and the sample size for each species of in the iris4 data frame and print iris6**

```{r echo=TRUE, warnings=FALSE, results=TRUE}
iris6 <- iris4 %>% 
#  select(Species)  # Why won't select() work?
group_by(Species) %>%
# use group_by() to creat a copy of data frame grouped by Species, then get summary 
# statistcs the same as above
summarise(Avg_Sepal.Length = mean(Sepal.Length),
         Avg_Sepal.Width = mean(Sepal.Width),
         SampleSize = n())
print(iris6)
```

**#8. In these exercises, you have successively modified different versions of the data frame iris1 iris2 iris3 iris4 iris5 iris6. At each stage, the output data frame from one operation serves as the input from the next.**  

* A more efficient way to do this is to use the pipe operator %>% from the tidyr package.   

* See if you can rework all of your previous statements (except for iris5) into an extended piping operation that uses iris as the input and generates irisFinal as the output.  

```{r echo=TRUE, warnings=FALSE, results=TRUE}
data("iris")

iris1 <-iris %>%
  filter(Species == "virginica" | Species == "versicolor",
        "Sepal.Length" > 6, 
        "Sepal.Width" > 2.5)  

iris2 <-iris1 %>%
  select(Species, Sepal.Length, Sepal.Width)  

iris3 <- iris2 %>% 
  arrange(desc(Sepal.Length))   

iris4 <- iris3 %>% 
  mutate(Sepal.Area = Sepal.Length * Sepal.Width)  

iris6 <- iris4 %>% 
  group_by(Species) %>%
  summarise(Avg_Sepal.Length = mean(Sepal.Length),
            Avg_Sepal.Width = mean(Sepal.Width),
            SampleSize = n())  

irisFinal <- iris6  
  
print(irisFinal)
```
* Or all at once

```{r echo=TRUE, warnings=FALSE, results=TRUE}
data("iris")
# Not able to automate the first step, but here are the rest of the operators all at once

irisFinal <-iris %>%
  filter(Species == "virginica" | Species == "versicolor",
         "Sepal.Length" > 6, 
         "Sepal.Width" > 2.5)  %>%
  select(Species, Sepal.Length, Sepal.Width)  %>%
  arrange(desc(Sepal.Length)) %>%
  mutate(Sepal.Area = Sepal.Length * Sepal.Width)  %>%
  group_by(Species) %>%
  summarise(Avg_Sepal.Length = mean(Sepal.Length),
            Avg_Sepal.Width = mean(Sepal.Width),
            SampleSize = n()) 

print(irisFinal)
```

**#9. Create a ‘longer’ data frame using the original iris data set with three columns named “Species”, “Measure”, “Value”.**  

* The column “Species” will retain the species names of the data set.  

* The column “Measure” will include whether the value corresponds to Sepal.Length, Sepal.Width, Petal.Length, or Petal.Width and the column “Value” will include the numerical values of those measurements.  

* I am not sure if we should be including all species from the data set, or just the two from the filtering in question #1. Below shown both ways.

```{r echo=TRUE, warnings=FALSE, results=TRUE}
#Pivot_longer lengthens the data, decreasing the number of columns, and increases the number of rows

#using pivot_longer() with all species
iris_long <- pivot_longer(iris, cols=c(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width), 
                        names_to="Measure", 
                        values_to = "Value")

head(iris_long)
tail(iris_long)


#using pivot_longer() with just "virginica" and "versicolor"
iris1<-filter(iris, Species == "virginica" | Species == "versicolor")

iris_long <- pivot_longer(iris1, cols=c(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width), 
                          names_to="Measure", 
                          values_to = "Value")
head(iris_long)
print(iris_long)
```

