---
title: "Homework #10"
author: "Nicole R Gorman"
date: "5 April 2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#For loops and randomization tests  

* Open libraries
```{r open libraries, message=FALSE}
library(tidyverse)
library(ggplot2)
library(MASS)
```

1. Using a for loop, write a function to calculate the number of zeroes in a numeric vector. Before entering the loop, set up a counter variable counter <- 0. Inside the loop, add 1 to counter each time you have a zero in the vector. Finally, use return(counter) for the output.

```{r echo=TRUE}

vec <- rbinom (n=5, size=1, 0.5)
counter <- 0
# create a vector w/ 0's and set up a counter variable

##########################################################################
# FUNCTION count_zeros
# count the number of zeros in a vector
# input: vector with zeros
# output: number of variables = 0
#----------------------------------------------------------------

count_zeros <- function(vec) {
 for (i in vec) {
   if(i==0) {
   counter <- counter + 1
}
   return(counter)
}
}

# end of count_zeros function
##########################################################################

count_zeros(vec)

```

2. Use subsetting instead of a loop to rewrite the function as a single line of code.

```{r echo=TRUE, results=TRUE}

vec <- rbinom (n=5, size=1, 0.5)
counter <- 0

##########################################################################
# FUNCTION count_zeros2
# count the number of zeros in a vector
# input: vector with zeros
# output: number of variables = 0
#----------------------------------------------------------------

count_zeros2 <- function(vec) {
   return(sum(vec==0))
}

# end of count_zeros2 function
##########################################################################

result <- count_zeros2(vec)
print (result)
# call function and print result
```



Write a function that takes as input two integers representing the number of rows and columns in a matrix. The output is a matrix of these dimensions in which each element is the product of the row number x the column number.

```{r echo=TRUE}

##########################################################################
# FUNCTION prod_matrix
# create matrix with elements as products of row and column
# input: number of rows and columns
# output: matrix of products
#----------------------------------------------------------------

prod_matrix <- function(nRow,nCol) {
   my_mat <- matrix(0, nrow = nRow, ncol = nCol)
  
  for (i in 1:nRow) {
    for (j in 1:nCol) {
       my_mat[i,j] <- i * j
  }
  
}
  return(my_mat)
}

# end of prod_matrix function
##########################################################################

my_matrix <- prod_matrix(3,4)
print (my_matrix)

```

Now let’s practice calling custom functions within a for loops. Use the code from previous lectures on loops and functions to complete the following steps:

Simulate a dataset with 3 groups of data, each group drawn from a distribution with a different mean. The final data frame should have 1 column for group and 1 column for the response variable.

```{r echo=TRUE}

# Set seed for reproducibility
set.seed(1)

# Number of observations per group
n_per_group <- 50

# Simulate data for each group
group1 <- rnorm(n_per_group, mean = 10, sd = 1)
group2 <- rnorm(n_per_group, mean = 15, sd = 1)
group3 <- rnorm(n_per_group, mean = 20, sd = 1)

# Combine data into a single dataframe
data <- data.frame(
  group = rep(c("Group1", "Group2", "Group3"), each = n_per_group),
  response = c(group1, group2, group3)
)

head(data)
tail(data)

```

Write a custom function that 1) reshuffles the response variable, and 2) calculates the mean of each group in the reshuffled data. Store the means in a vector of length 3.

```{r echo=TRUE}
##########################################################################
# FUNCTION means
# reshuffle response variable, calculate the mean of each group, store the means in a vector of length 3.
# input: dataset with 3 groups
# output: three group means as vector
#----------------------------------------------------------------

data2 <- function(data) {
   res2 <- sample(response)
  
  mean
  
}
  return(my_mat)
}

# end of prod_matrix function
##########################################################################

my_matrix <- prod_matrix(3,4)
print (my_matrix)

```


Write a custom function that 1) reshuffles the response variable, and 2) calculates the mean of each group in the reshuffled data. Store the means in a vector of length 3.

Use a for loop to repeat the function in b 100 times. Store the results in a data frame that has 1 column indicating the replicate number and 1 column for each new group mean, for a total of 4 columns.

Use qplot() to create a histogram of the means for each reshuffled group. Or, if you want a challenge, use ggplot() to overlay all 3 histograms in the same figure. How do the distributions of reshuffled means compare to the original means?


```
#Let's talk about tea
# Is milk added first or not?
# H0 vs HA
# Blind test
# can reshuffle the data to repeat the experiment rather than actually redoing the experiment
```

```
# question 3 is going to require double for loop
```
