---
title: "Homework11"
author: "Nicole R Gorman"
date: "2024-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Batch Processing 

* Open libraries
```{r open libraries, message=FALSE}
library(log4r)
library(TeachingDemos)
library(tidyverse)
library(pracma)
library(ggmosaic)
```

### Question #1.
```{r echo = TRUE, results = TRUE}

## Download dataset ----
## Set working directory ----
dir <- "~/Documents/UVM_Coursework/ComputationalBiology/Original Data"
```

### Question #2.
#### Within each year’s folder, you will only be using a file from each year labeled “countdata” in its title. Using for loops, iterate through each year’s folders to gather the file names of these “countdata” .csv files.

```{r echo = TRUE, results = FALSE}

# I don't think I have this set up correctly, and I am not even sure if I need it, but I was not able to set a working directory in the for loop without throwing an error
# I tried to use the recursive argument, but I am not sure I am using it properly
# I used `full.names` to keep track of the pathname because that is what I was struggling with

# iterate through each folder-go through each folder-go to levels down each time 
# pattern ="countdata"

filenames <- list.files(dir, pattern ="countdata", recursive = TRUE, full.names = TRUE) 
print(filenames)

# Read in the csv files
# data_list contains dataframes
data_list <- lapply(filenames, read.csv) 

# use a loop to pull out files
# create empty object, then fill it in the loop
my_files <- c()

for (i in seq_along(filenames)) {
   my_files <- c(my_files, filenames[i])
 }
 
 print(my_files)
```

### Question #3. 
#### Step 1: Create Pseudo-code

```{r echo = TRUE, results = TRUE}
# Starting with pseudo-code, generate functions for 1) Cleaning the data for any empty/missing cases 2) Extract the year from each file name, 3) Calculate Abundance for each year (Total number of individuals found), 4) Calculate Species Richness for each year(Number of unique species found)

### Step 2: Write function for each list item ----
##### 1) Cleaning the data ----
# for any empty/missing cases 

##########################################################################
# FUNCTION clean_data
# read in data, omit NAs
# input: data set 
# output: cleaned data set 
#__________________________________________________________________________

clean_data <- function(data) {
  cleaned_data_list <- list() # create empty list
  
  for (i in seq_along(filenames)) {
    data <- read.csv(filenames[i], row.names = NULL) # Read CSV file
    cleaned <- data[complete.cases(data$scientificName), ]  # remove NAs to clean data
  
# I know I need to store cleaned data in the list, not sure how to do this
# Cleaned_data_list is supposed to contain the cleaned data for each file
#    cleaned_data_list[[i]] <- cleaned_data
}
    
    return(cleaned_data_list)
  }

# end of clean_data function
##########################################################################

cleaned_data <- clean_data(data)
head (cleaned_data)

##### 2) Extract the year ----
# from each file name

##########################################################################
# FUNCTION extract_year
# Extract year from each file name, "_countdata" followed by date "YYYY-MM", replaces filenames w/ only year
# input: data set 
# output: list of years
#__________________________________________________________________________

extract_year <- function(filename) {
 
  years <- gsub(".*_countdata\\.(\\d{4})-\\d{2}.*\\.csv", "\\1", filename)  # Extract the year from each filename
  years <- as.numeric(years)   # Convert the years to numeric
  years_list <- as.list(years)   # Convert the numeric vector to a list
  print(years_list) # Print the extracted years
  
  return(years)
}

# end of clean_data function
##########################################################################

#results <- extract_year()
#print(results)

##### 3) Calculate Abundance ----
# for each year (Total number of individuals found), 

##########################################################################
# FUNCTION abundance
# Read count data from each file and calculate abundance for each year (total number of individuals found)
# input: data set, years 
# output: abundance for each year
#__________________________________________________________________________

abundance <- function(filename) {
  year_abundance <- numeric()  # Create an empty numeric vector to store abundance for each year
  count_data <- read.csv(filename) # Read count data from file
  year <- extract_year(filename)  # Extract year from filename
  year_abundance[year] <- sum(count_data$Count) # Calculate abundance
  return(year_abundance)
}

# end of abundance function
##########################################################################

#results <- abundance("NEON.D01.BART.DP1.10003.001.brd_countdata.2022-06.basic.20231229T053256Z.csv")
#tail (results)

##### 4) Calculate Species Richness ----
# for each year (Number of unique species found)

##########################################################################
# FUNCTION calculate_richness
# Calculate species richness for each year (number of species found)
# input: data set 
# output: species richness
#__________________________________________________________________________

calculate_richness <- function(filename) {
  count_data <- read.csv(filename) # Read count data from file
  year <- extract_year(filename) # Extract year from filename
  richness <- length(unique(count_data$Species)) # Calculate richness
  return(richness)
}

# end of richness function
##########################################################################

#results <- calculate_richness("NEON.D01.BART.DP1.10003.001.brd_countdata.2022-06.basic.20231229T053256Z.csv")
#print (results)
```

### Question #4. 
#### Create empty data frame to hold the above summary statistics
```{r echo = TRUE, results = TRUE}

# create empty data frame with 4 columns, one for the file name, one for abundance, one for species richness, and one for year.
# Not clear about whether I need the stringsAsFactors bit

sum_stats <- data.frame(
  filename = character(),
  abundance = numeric(),
  species_richness = numeric(),
  year = integer(),
  stringsAsFactors = FALSE
)

str(sum_stats)
```

### Question #5. 
#### Step 3: Create function templates as batch operation 

```{r echo = TRUE, results = TRUE}

### Step 4: Source all function templates as a batch operation 
#### Using a for loop, run your created functions as a batch process for each folder, changing the working directory as necessary to read in the correct files, calculating summary statistics with your created functions, and then writing them out into your summary statistics data frame.

# Get the list of subdirectories within the main directory
subdirs <- list.dirs(dir, recursive = TRUE)

# Iterate over each subdirectory
for (subdir in subdirs) {
  setwd(subdir) # Set working directory to the current subdirectory
  filenames <- list.files(pattern = "countdata", full.names = TRUE) # Get the filenames
  subdir_summary <- list() # Create an empty list 
  
# Iterate over each file in the subdirectory  
for (filename in filenames) {
    cleaned_data <- clean_data(read.csv(filename))   # Clean the data
    year <- extract_year(filename)   # Extract the year from the filename
    file_abundance <- sum(cleaned_data$Count)  # Calculate abundance for the current file
    file_richness <- calculate_richness(filename) # Calculate species richness for the current file
    
    subdir_summary[[filename]] <- list( 
      filename = filename,
      abundance = file_abundance,
      species_richness = file_richness,
      year = year
    )
  }

subdir_summary_df <- do.call(rbind, subdir_summary) # Combine summary statistics f
sum_stats <- rbind(sum_stats, subdir_summary_df) # Append the subdir_summary_df df 
}
print(sum_stats) # Print the summary statistics data frame

```