---
title: "Homework11"
author: "Nicole R Gorman"
date: "2024-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Batch Processing** 

```{r open libraries, message=FALSE}

# install libraries
library(log4r)
library(TeachingDemos)
library(tidyverse)
library(pracma)
library(ggmosaic)
```

#### Question #1
```{r echo = TRUE, results = TRUE}

## Download dataset ----
## Set working directory ----
dir <- "~/Documents/UVM_Coursework/ComputationalBiology/Original Data"
```

#### Question #2
##### Within each year’s folder, you will only be using a file from each year labeled “countdata” in its title. Using for loops, iterate through each year’s folders to gather the file names of these “countdata” .csv files.

```{r echo = TRUE, results = FALSE}

# I struggled with this step. I was able to get the file names but had trouble going down into each individual directory. I don't think I have it set up correctly, and I am not sure if I need all the components of the code. It threw me off because I was not able to set a working directory in the for loop without throwing an error.

# I ended up trying to use the recursive argument, but I am not sure I am using it properly
# I used `full.names` to keep track of the pathname because it seems like that is what I was struggling with and I thought it would be helpful if I could see the path

# iterate through each folder-go through each folder-go to levels down each time 
# pattern ="countdata"

filenames <- list.files(dir, pattern ="countdata", recursive = TRUE, full.names = TRUE) 
print(filenames)

# Read in the csv files
# data_list contains dataframes
data_list <- lapply(filenames, read.csv) 


# create empty object, then fill it in the loop
my_files <- c()

# use a loop to pull out files
# fill the object I created above
# I am still a bit unclear how to properly write a loop

for (i in seq_along(filenames)) {
  my_files <- c(my_files, filenames[i])
 }
 
 print(my_files)
```

#### Question #3 
##### Starting with pseudo-code, generate functions for 1) Cleaning the data for any empty/missing cases 2) Extract the year from each file name, 3) Calculate Abundance for each year (Total number of individuals found), 4) Calculate Species Richness for each year(Number of unique species found)

```{r echo = TRUE, results = TRUE}

# Step 1: Create Pseudo-code
# Step 2: Write function for each list item 

# 1) Cleaning the data for any empty/missing cases 

##########################################################################
# FUNCTION clean_data
# create empty list, read in data, omit NAs, return data list with NAs removed
# input: data set 
# output: cleaned data set 
#__________________________________________________________________________

clean_data <- function(data) {
  cleaned_data_list <- list() # create empty list
  
  for (i in seq_along(filenames)) {
    data <- read.csv(filenames[i], row.names = NULL) # Read CSV file
    cleaned <- data[complete.cases(data$scientificName), ]  # remove NAs to clean data
  
# I know I need to store cleaned data in the list, but I am not sure how to do this
# Cleaned_data_list is supposed to contain the cleaned data for each file

# tried several iterations, maybe something like this should be in the loop
# cleaned_data_list[[i]] <- cleaned_data
}
    
    return(cleaned_data_list)
  }

# end of clean_data function
##########################################################################

# check the function and take a peek at the new dataset
cleaned_data <- clean_data(data)
head (cleaned_data)

# 2) Extract the year from each file name

##########################################################################
# FUNCTION extract_year
# Extract year from each file name w/ "countdata" using reg expressions, convert years, make a vector, print list
# input: data set 
# output: list of years
#__________________________________________________________________________

extract_year <- function(filename) {
 
  years <- gsub(".*_countdata\\.(\\d{4})-\\d{2}.*\\.csv", "\\1", filename)  # Extract  year from each filename
  years <- as.numeric(years)   # Convert the years to numeric vector
  years_list <- as.list(years)   # Convert the numeric vector to a list
  print(years_list) # Print  extracted years
  
  return(years)
}

# end of clean_data function
##########################################################################

# Rmd file would not run with the following two lines of code
# I may be using an improper argument for the new function 
# results <- extract_year()
# print(results)

#3) Calculate Abundance for each year (Total number of individuals found), 

##########################################################################
# FUNCTION abundance
# create empty vector, read in files, extract year, calculate abundance, return abundance
# input: data set, years 
# output: abundance for each year
#__________________________________________________________________________

abundance <- function(filename) {
  year_abundance <- numeric()  # Create empty numeric vector to store abundance for each year
  count_data <- read.csv(filename) # Read count data from file
  year <- extract_year(filename)  # Extract year from filename
  year_abundance[year] <- sum(count_data$Count) # Calculate abundance
  return(year_abundance)
}

# end of abundance function
##########################################################################

# Rmd file would not run with the following two lines of code calling the new function
# I may be using an improper argument for the new function 
# results <- abundance("NEON.D01.BART.DP1.10003.001.brd_countdata.2022-06.basic.20231229T053256Z.csv")
# tail (results)

# 4) Calculate Species Richness for each year (Number of unique species found)

##########################################################################
# FUNCTION calculate_richness
# read in data, extract year, calculate richness
# input: data set 
# output: species richness
#__________________________________________________________________________

calculate_richness <- function(filename) {
  count_data <- read.csv(filename) # Read count data from file
  year <- extract_year(filename) # Extract year from filename
  richness <- length(unique(count_data$Species)) # Calculate richness
  return(richness)
}

# end of richness function
##########################################################################

# Rmd file would not run with the following two lines of code calling the new function
# I may be using an improper argument for the new function 
# results <- calculate_richness("NEON.D01.BART.DP1.10003.001.brd_countdata.2022-06.basic.20231229T053256Z.csv")
# print (results)
```

#### Question #4 
##### Create empty data frame to hold the above summary statistics.
```{r echo = TRUE, results = TRUE}

# create empty data frame with 4 columns, one for the file name, one for abundance, one for species richness, and one for year.
# Not clear about whether I need the stringsAsFactors bit

sum_stats <- data.frame(
  filename = character(),
  abundance = numeric(),
  species_richness = numeric(),
  year = integer(),
  stringsAsFactors = FALSE
)

str(sum_stats)
```

#### Question #5 
##### Create function templates as batch operation 
##### Using a for loop, run your created functions as a batch process for each folder, changing the working directory as necessary to read in the correct files, calculating summary statistics with your created functions, and then writing them out into your summary statistics data frame.
```{r echo = TRUE, results = TRUE}

# Step 3: Create function templates as batch operation
# Step 4: Source all function templates as a batch operation 


# This in disorganized and unsuccessful. As I am sure you can see, I am pretty confused about how to structure this code. At a vary basic level, I can't figure out how to navigate the subdirectories. It seems like I should be able to use setwd, but it throws an error inside a for loop. 

# Get the list of subdirectories within the main directory
subdirs <- list.dirs(dir, recursive = TRUE)

# Iterate over each subdirectory
for (subdir in subdirs) {
  setwd(subdir) # Set working directory to the current subdirectory
  filenames <- list.files(pattern = "countdata", full.names = TRUE) # Get the filenames
  subdir_summary <- list() # Create an empty list 
  
# Iterate over each file in the subdirectory  
for (filename in filenames) {
    cleaned_data <- clean_data(read.csv(filename))   # Clean the data
    year <- extract_year(filename)   # Extract the year from the filename
    file_abundance <- sum(cleaned_data$Count)  # Calculate abundance for the current file
    file_richness <- calculate_richness(filename) # Calculate species richness for the current file
    
    subdir_summary[[filename]] <- list( 
      filename = filename,
      abundance = file_abundance,
      species_richness = file_richness,
      year = year
    )
  }

subdir_summary_df <- do.call(rbind, subdir_summary) # Combine summary statistics f
sum_stats <- rbind(sum_stats, subdir_summary_df) # Append the subdir_summary_df df 
}
print(sum_stats) # Print the summary statistics data frame

```