---
title: "Working with Markdown"
author: "Nicole R Gorman"
date: "2024-01-23"
output: 
  html_document: 
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For markdown, use asterisks to *indicate italic text*.

```{r}

#first chunk of code in rmd file
print(pi) 
pi + 3

```

Double line return for a new chunk of text like this.

Nicole R. Gorman  
Department of Plant Biology  
University of Vermont  

**#  Practice with RMarkdown code**  
**##  Different language to change formatting**

# Hash marks for headers

## More hashes for progressively

## smaller

### and smaller

#### headers!!

**bold face this way**

>Offset quote which embeds and looks really nice when you have extended material that you would like to see.

Back to plain text

* first list item
* second list item
   - spacing for indents  

^superscript^ but also a ~superscript~ like this

[set up a link with text](website address)

First Header  |  Second Header
------------  |  -------------
Content Cell  |  Content Cell
Content Cell  |  Content Cell

'''
Fencing sets boundaries and everything inside the fencing *is shown as literal text* without rendering.
'''

$$a + b =c$$

We can also use Latex in the line with a single dollar sign $a + b=c$ like this.

##subscripts in LaTex

$$H_0 = Z_{a + b}$$




$$S=cA^z_1+ z_(Z + x)$$

$$\alpha = \frac{\best}{\delta + \gamma_x}$$
$$z = \sum_{i=1}^X{K}$$


### What if we need a backslash??

$$\backslash \alpha \le b \ \backslash$$

### Sometimes we want to mix math with plain text?


$$P(Occurance Of Species A) = Z$$
$$P(\mbox{Occurance Of Species A}) = Z$$

East to embed Latex in line $\pi + 3$ like so we can also use markdown fencing in line with a single backtick `pi + 3` like this. We can even execute r code inside markdown text `r pi +3`


# Rendering in plain text

We can run and execute R code within the R Markdown file. We can add code insode the fencing.
  
  
```{r eval=TRUE, echo=TRUE}

# use the eval statement to toggle 

#{r eval=TRUE, echo=FALSE}
# Commented this bc notr working

# Use comments extensively in all of your coding!
Pred <-seq(1,10)   #make a vector of integers from 1 to 10
Res <-runif(10)    #generate 10 draws from a random uniform (0,1) distribution

# Print random numbers
print(Res)

plot(Res,Pred)

```
